\section{Related Work}\label{sec:related_work}

\paragraph{Benchmark contamination detection.}
Contamination detection methods span a wide range of information requirements.
Corpus-search methods match benchmark items directly against pretraining data \citep{dodge2021, magar2022}, but require full corpus access unavailable for closed models.
Membership inference attacks (MIA) use token-level log-probabilities to distinguish training from unseen examples \citep{minkprob2024, deng2024}.
Generation-based probes elicit benchmark content via prompting \citep{golchin2024, golchin2024quiz, oren2024}.
Performance-based methods compare accuracy on target and reference sets \citep{constat2024, pacost2024, codec2026}, requiring multiple evaluations or in-context learning passes.
\citet{zhou2023benchmark} and \citet{sainz2023} argue for systematic contamination measurement, \citet{jacovi2023} propose mitigation strategies, and \citet{bordt2025forget} show that contamination can be ``forgotten'' under sufficient continued training.
\citet{han2026quantifying} quantify contamination in psychometric evaluations of LLMs, demonstrating that memorization effects extend beyond standard benchmarks.
\citet{kattamuri2025radar} take a mechanistic interpretability approach, using activation patterns to distinguish recall-based from reasoning-based responses; our work instead operates purely on behavioral response patterns without access to model internals.
\citet{fu2025contamination} systematically evaluate detection method assumptions, finding that most methods degrade under realistic conditions; our work contributes a psychometric perspective to this ongoing evaluation.

A critical limitation unites all token-level methods: \citet{wang2025fragility} show that RL post-training renders MIA-based detectors no better than random across 10 methods and 4 benchmarks, because RL reshapes token distributions while largely preserving binary correctness patterns.
\citet{selfcritique2026} and \citet{zhang2026rlvr} propose RL-specific solutions but require generation-level access.
Our approach requires no access to model weights, logits, or training data---only binary response patterns and externally calibrated item parameters from PSN-IRT \citep{psnirt2026}.

\paragraph{IRT for LLM evaluation.}
Item Response Theory has seen growing adoption in NLP and ML evaluation.
\citet{martinezplumed2019} applied IRT to analyze machine learning classifiers at the instance level, establishing that item difficulty and discrimination parameters capture meaningful variation across models.
\citet{lalor2019} proposed estimating IRT parameters from DNN-generated response patterns, removing the need for human annotators.
\citet{rodriguez2021} demonstrated that IRT-informed item weighting improves leaderboard rankings over raw accuracy.
\citet{polo2024tinybenchmarks} used IRT-based item selection to reduce benchmark sizes by 100$\times$ while preserving model rankings, and \citet{atlas2025} developed adaptive testing procedures for efficient LLM evaluation.
Most directly relevant, \citet{psnirt2026} calibrated 4PL item parameters for over 13,000 models across 11 benchmarks, providing the item bank our framework builds on.
All of these works use IRT for ability estimation or benchmark efficiency; none apply \emph{person-fit} statistics for contamination detection.

\paragraph{Person-fit statistics in psychometrics.}
Person-fit analysis has a long history in educational measurement for detecting aberrant test-taking behavior \citep[see][for a textbook treatment]{embretson2000}.
\citet{levine1988} formalized optimal appropriateness measurement; \citet{drasgow1985} introduced $\ell_z$, a standardized log-likelihood residual; \citet{snijders2001} derived the corrected $\ell_z^*$ that accounts for ability estimation uncertainty.
\citet{karabatsos2003} compared 36 person-fit statistics and found $\ell_z$ among the most powerful for detecting cheating; \citet{sinharay2017personfit} revisited these simulations and confirmed comparable power between parametric and nonparametric statistics.
We select $\ell_z^*$ for its theoretical grounding and HT \citep{vanderflier1982} as a nonparametric complement, following their top-ranked performance in that comparison.
\citet{reise1990} and \citet{meijer2001} provide systematic evaluations of person-fit methodology.
\citet{sinharay2017} showed that $\ell_z^*$ is particularly effective when the set of compromised items is known, a setting analogous to benchmark contamination where memorized items constitute the compromised set.
Our work transfers this psychometric toolkit to LLM evaluation, treating each model as an examinee whose item-level response pattern may reveal memorization.
