\documentclass[11pt]{article}
\usepackage[review]{acl}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{enumitem}

\newcommand{\irtfit}{\textsc{IRT-FIT}}

\graphicspath{{../figures/paper/}}

\title{What Can IRT Person-Fit Statistics Reveal About SFT-Stage Benchmark Contamination?}

\author{Anonymous}

\begin{document}
\maketitle

\begin{abstract}
\input{sections/abstract}
\end{abstract}

\input{sections/introduction}
\input{sections/related_work}
\input{sections/method}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

\input{sections/limitations}

\section*{Ethical Considerations}
This work demonstrates how SFT-stage benchmark contamination can be detected and characterized using IRT person-fit statistics.
While the contamination methodology we employ (fine-tuning on benchmark items) is well-established in prior work, we acknowledge the dual-use nature of systematic contamination studies: the same controlled framework could inform adversarial contamination strategies.
We believe the detection and diagnostic contributions outweigh this risk, as they support benchmark integrity and more reliable model evaluation.
All experiments use publicly available benchmarks under permissive licenses and open-weight models; no private data or human subjects are involved.

\section*{Reproducibility Statement}
Code and response data are available in an anonymized repository.
Bootstrap confidence intervals use the percentile method with 10,000 resamples.
All three model families use clean baseline seeds 0--9 ($n{=}10$ per family).
Pre-calibrated 4PL item parameters are from the publicly available PSN-IRT item bank \citep{psnirt2026}.
Training used PyTorch 2.8.0, Transformers 5.8.1, PEFT 0.19.1, and BitsAndBytes 0.49.2.
Both benchmarks are publicly available under permissive licenses (MMLU: MIT; ARC-Challenge: CC-BY-SA).

\bibliography{references}

\appendix
\input{sections/appendix}

\end{document}
